A data-driven model of nucleosynthesis with chemical tagging in latent space

### 2019-05-20 article/mcfa.tex
moving furniture and cleaning house

\documentclass[twocolumn]{aastex62}
\usepackage{bm}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{color}
\usepackage{comment}
%\usepackage{layouts}

@@

\newcommand{\NumData}{N}
\newcommand{\NumDimensions}{D}
\newcommand{\numdata}{n}
\newcommand{\numdimensions}{d}
\newcommand{\NumLatentFactors}{J}
\newcommand{\numlatentfactors}{j}
\newcommand{\NumComponents}{K}
\newcommand{\numcomponents}{k}
\newcommand{\likelihood}{\mathcal{L}}

@@

\begin{abstract}
Chemical tagging seeks to identify unique star formation sites from
present-day stellar abundances.
Previous techniques have treated chemical abundances as being 
statistically independent, despite the nucleosynthesis literature showing that
most elements can be produced by more than one nucleosynthetic pathway.
In this work we introduce a data-driven model of nucleosynthesis 
where a set of latent factors (e.g., nucleosynthetic yields) contribute
to all stars with different scores, and clustering (e.g., chemical tagging) 
is represented by a mixture model of multivariate normal distributions
in a lower-dimensional latent space.
We use an exact method to simultaneously estimate the factor scores for
each star, the relative weighting (e.g., partial assignment) of each star
to each cluster, and to infer the latent factors that are common to all stars.
We use an information-theoretic Bayesian principle to estimate the number of
latent factors and clusters.
We perform experiments on generated and real data, including literature chemical abundances
of metal-poor stars, and those from the second \Galah\ data release. 
We find that \todo{five} latent factors are preferred to explain the detailed 
abundances of metal-poor stars, and \todo{eight} are preferred given \todo{X} abundances
in \Galah\ data. In both experiments we recover latent factors that 
are representative of mean yields expected from the rapid- and slow-neutron
capture process, as well as nucleosynthesis in Type II supernovae. 
We find \todo{two} clusters are preferred in the \Galah\ data of 16 abundances
and 1,136 stars, which are representative of the [$\alpha$/Fe]-poor and 
[$\alpha$/Fe]-rich populations in the Milky Way disk. 
We expect more components
would be preferred given complete abundance inventory of a larger sample of stars.
We release accompanying software that scales well with the available data,
allowing model parameters to be optimised in a few minutes even given $\sim$$10^{7.5}$ 
abundance measurements.
%We release software that scales \todo{as $\mathcal{O}(XX)$}, where $X$ is \todo{what},
%enabling model parameter estimation in minutes given $\sim$$10^7$
%chemical abundance measurements.
\end{abstract}

@@

The Fisher information for $\factorloads$ \todo{is hard to compute. I have 
the first derivative from the log likelihood function, but the second
derivative is (surprisingly) causing some trouble.}
Thus the message length to encode $\factorloads$ is given by
\begin{eqnarray}
	I(\factorloads) &=& -\log\left(\frac{p(\factorloads)}{\sqrt{|\mathcal{F}(\factorloads)|}}\right) \nonumber \\
I(\factorloads)	&=& \frac{1}{2}\textrm{Tr}(\textrm{Cov}(\factorloads\transpose)^{-1}\factorloads\transpose\factorloads) - \frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)\log{|\factorloads\transpose\factorloads|} \nonumber \\
	&& \cdots \,\, + \frac{1}{2}\NumDimensions\NumLatentFactors\log{2} + \frac{1}{2}\NumDimensions\log{|\textrm{Cov(\factorloads)}|} - \Gamma\left(\frac{\NumDimensions}{2}\right) \nonumber \\
	&& \cdots \todo{+\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}} \quad . \label{eq:prior_L}
\end{eqnarray}

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-log-xfe-offset-gridsearch-ll.pdf}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-log-xfe-offset-gridsearch-bic.pdf}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-log-xfe-offset-gridsearch-mml.pdf}
    \caption{The top panel shows the negative log likelihood 
			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
			 evaluated at each combination of latent factors 
			 $\NumLatentFactors$ and number of clusters 
			 $\NumComponents$ using the \citet{Barklem:2005}
			 data in Experiment~2.  The middle panel shows 
			 the BIC (Eq.~\ref{eq:bic}) for those 
			 combinations, and the lower panel shows the 
			 message length. The white marker indicates the
			 lowest value in each panel, showing the
			 preferred number of latent factors and components.}
    \label{fig:exp2-gridsearch-contours}
\end{figure}

@@

\begin{figure*}
	\includegraphics[width=\textwidth]{experiments/exp2-log-xfe-offset-latent-factors-visualize-abs.pdf}
	\caption{Latent factors found in Experiment~2 using the \citet{Barklem:2005}
			 data. The panels on the left show the absolute entries for each factor load. 
			 On the right we show the absolute fractional contributions to
			 each element, illustrating the loads that contribute most to each element.}
    \label{fig:exp2-factor-loads}
\end{figure*}

@@

\begin{figure}
	%\includegraphics[width=0.45\textwidth]{experiments/exp2-log-xfe-offset-specific-scatter.pdf}
	\includegraphics[width=0.5\textwidth]{../experiments/exp2-specific-scatter.pdf}
	\caption{Specific scatter $\sqrt{\specificvariance}$ remaining in the \citet{Barklem:2005} data used in
			 Experiment~2 after accounting for the contributions by all latent
			 factors.}
    \label{fig:exp2-specific-scatter}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-gridsearch-ll.pdf}
	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-gridsearch-bic.pdf}
	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-gridsearch-mml.pdf}
    \caption{The top panel shows the negative log likelihood 
			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
			 evaluated at each combination of latent factors 
			 $\NumLatentFactors$ and number of clusters 
			 $\NumComponents$ using \Galah\ data in
			 Experiment~3.  The middle panel shows 
			 the BIC for those combinations, and the lower panel shows the 
			 message length. The white marker indicates the
			 lowest value in each panel, showing the
			 preferred number of latent factors and components.}
    \label{fig:experiment-9-gridsearch}
\end{figure}

\begin{figure*}
	\includegraphics[width=\textwidth]{experiments/exp9-log-xfe-offset-latent-factors-visualize-abs.pdf}
	\caption{Latent factors inferred in Experiment~3 using the \Galah\
			 data. The panels on the left show the absolute entries for each
			 factor load. On the right we show the absolute fractional contributions
			 to each element, illustrating the loads that contribute most to each
			 element.}
    \label{fig:exp9-factor-loads}
\end{figure*}

@@

\begin{comment}
\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp4-abundance-counts.pdf}
	\caption{The number of abundance measurements in the second
		     \Galah\ data release \citep{Buder:2018} where \texttt{flag\_cannon = 0}
		     and \texttt{flag\_<element> = 0}. The black line indicates the number
		     of reported measurements for a single element, and the blue line
		     indicates the number of stars where all elements leftward (inclusive)
		     are reported. The dashed vertical line indicates the top 18 reported
		     elemental abundances, which we use for Experiment~4.}
    \label{fig:exp4-abundance-counts} 
\end{figure}
\end{comment}

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-specific-scatter.pdf}
	\caption{Specific scatter $\sqrt{\specificvariance}$ remaining in the \Galah\ data \citep{Buder:2018}
			 in Experiment~3 after accounting for the contributions by all
			 latent factors.}
    \label{fig:exp9-specific-scatter}
\end{figure}

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-cluster.pdf}
	\caption{Abundance ratios of [Fe/H] and [Mg/Fe] for all stars in Experiment~3,
			 colored by their probability of being associated with the first (of two)
			 components. This illustrates how the clustering in latent space is projected
			 to real space, and indicates that the two components identified are largely
			 consistent with being stars with low- and high-$\alpha$ element abundance
			 ratios.}
    \label{fig:exp9-cluster}
\end{figure}

>>>>

\documentclass[twocolumn]{aastex62}
\usepackage{bm}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{color}
\usepackage{comment}
\usepackage{layouts}

@@

\newcommand{\NumData}{N}
\newcommand{\NumDimensions}{D}
\newcommand{\numdata}{n}
\newcommand{\numdimensions}{d}
\newcommand{\NumLatentFactors}{J}
\newcommand{\numlatentfactors}{j}
\newcommand{\NumComponents}{K}
\newcommand{\numcomponents}{k}
\newcommand{\likelihood}{\mathcal{L}}
\newcommand{\ExperimentHash}{96ff8}

@@

\begin{abstract}
Chemical tagging seeks to identify unique star formation sites from
present-day stellar abundances.
Previous techniques have treated the abundances of each element as being
statistically independent, despite the nucleosynthesis literature showing that
most elements can be produced by more than one pathway.
In this work we introduce a data-driven model of nucleosynthesis 
where a set of latent factors (e.g., nucleosynthetic yields) contribute
to all stars with different scores, and clustering (e.g., chemical tagging) 
is represented by a mixture model of multivariate gaussians
in a lower-dimensional latent space.
We use an exact method to simultaneously estimate the factor scores for
each star, the partial assignment of each star
to each cluster, and the latent factors common to all stars.
We use an information-theoretic Bayesian principle to estimate the number of
latent factors and clusters.
We perform experiments on generated and real data, including literature chemical abundances
of metal-poor stars, and those from the second \Galah\ data release. 
We find that \todo{five} latent factors are preferred to explain the detailed 
abundances of metal-poor stars, and \todo{eight} are preferred given \todo{X} abundances
in \Galah\ data. In both experiments we recover latent factors that 
are representative of mean yields expected from the rapid- and slow-neutron
capture process, as well as nucleosynthesis in Type II supernovae. 
We find \todo{two} clusters are preferred in the \Galah\ data of 16 abundances
and 1,136 stars, which are representative of the [$\alpha$/Fe]-poor and 
[$\alpha$/Fe]-rich populations in the Milky Way disk. 
We expect more components
would be preferred given complete abundance inventory of a larger sample of stars.
We release accompanying software that scales well with the available data,
allowing for model parameters to be optimised in seconds given a fixed number
of latent factors, components, and $\sim$$10^{7.5}$ abundance measurements.
%We release software that scales \todo{as $\mathcal{O}(XX)$}, where $X$ is \todo{what},
%enabling model parameter estimation in minutes given $\sim$$10^7$
%chemical abundance measurements.
\end{abstract}

@@

\printinunitsof{in}\prntlen{\textheight}
and 
\printinunitsof{in}\prntlen{\textwidth}

%\todo{Fisher:}
%\begin{widetext}
%\begin{eqnarray}
%	\log\likelihood(\vecdata|\vec\Psi_k) = -\frac{1}{2}(\vecdata - \factorloads\vec\xi_k)\transpose(\factorloads\omega_k\factorloads\transpose + \diag{D})^{-1}) - \frac{1}{2}\log|(\factorloads\omega_k\factorloads\transpose + \diag{D})| -\frac{1}{2}\log{2\pi}
%	\log\likelihood(\vecdata|\vec\Psi) &=& -\frac{1}{2}(\vecdata - \vec{M})\transpose\vec\Sigma^{-1}(\vecdata - \vec{M}) - \frac{1}{2}\log|\vec\Sigma| - \frac{1}{2}\log{2\pi} \quad \mathrm{where} \\
%	\vec{M} &=& \factorloads\scoremeans \\
%	\vec\Sigma &=& \factorloads\scorecovs\factorloads\transpose + \diag{\specificvariance}
%\end{eqnarray}
%\end{widetext}

The Fisher information for $\factorloads$ \todo{needs to be checked.}
Thus the message length to encode $\factorloads$ is given by
\begin{eqnarray}
	I(\factorloads) &=& -\log\left(\frac{p(\factorloads)}{\sqrt{|\mathcal{F}(\factorloads)|}}\right) \nonumber \\
I(\factorloads)	&=& \frac{1}{2}\textrm{Tr}(\textrm{Cov}(\factorloads\transpose)^{-1}\factorloads\transpose\factorloads) - \frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)\log{|\factorloads\transpose\factorloads|} \nonumber \\
	&& \cdots \,\, + \frac{1}{2}\NumDimensions\NumLatentFactors\log{2} + \frac{1}{2}\NumDimensions\log{|\textrm{Cov(\factorloads)}|} - \Gamma\left(\frac{\NumDimensions}{2}\right) \nonumber \\
	&& \cdots \todo{+\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}} \quad . \label{eq:prior_L}
\end{eqnarray}

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-\ExperimentHash-gridsearch-ll.pdf}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-\ExperimentHash-gridsearch-bic.pdf}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-\ExperimentHash-gridsearch-mml.pdf}
    \caption{The top panel shows the negative log likelihood 
			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
			 evaluated at each combination of latent factors 
			 $\NumLatentFactors$ and number of clusters 
			 $\NumComponents$ using the \citet{Barklem:2005}
			 data in Experiment~2.  The middle panel shows 
			 the BIC (Eq.~\ref{eq:bic}) for those 
			 combinations, and the lower panel shows the 
			 message length. The white marker indicates the
			 lowest value in each panel, showing the
			 preferred number of latent factors and components.}
    \label{fig:exp2-gridsearch-contours}
\end{figure}

@@

\begin{figure*}
	\includegraphics[width=\textwidth]{experiments/exp2-\ExperimentHash-latent-factors-and-contributions.pdf}
	\caption{Latent factors found in Experiment~2 using the \citet{Barklem:2005}
			 data. \todo{The panels on the left show the absolute entries for each factor load.} 
			 On the right we show the absolute fractional contributions to
			 each element, illustrating the loads that contribute most to each element.}
    \label{fig:exp2-factor-loads}
\end{figure*}

@@

\begin{figure}
	\includegraphics[width=0.5\textwidth]{../experiments/exp2-\ExperimentHash-specific-scatter.pdf}
	\caption{Specific scatter $\sqrt{\specificvariance}$ remaining in the \citet{Barklem:2005} data used in
			 Experiment~2 after accounting for the contributions by all latent
			 factors.}
    \label{fig:exp2-specific-scatter}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp3-\ExperimentHash-gridsearch-ll.pdf}
	\includegraphics[width=0.45\textwidth]{experiments/exp3-\ExperimentHash-gridsearch-bic.pdf}
	\includegraphics[width=0.45\textwidth]{experiments/exp3-\ExperimentHash-gridsearch-mml.pdf}
    \caption{The top panel shows the negative log likelihood 
			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
			 evaluated at each combination of latent factors 
			 $\NumLatentFactors$ and number of clusters 
			 $\NumComponents$ using \Galah\ data in
			 Experiment~3.  The middle panel shows 
			 the BIC for those combinations, and the lower panel shows the 
			 message length. The white marker indicates the
			 lowest value in each panel, showing the
			 preferred number of latent factors and components.}
    \label{fig:exp3-gridsearch}
\end{figure}

\begin{figure*}
	\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-latent-factors-and-contributions.pdf}
	\caption{Latent factors inferred in Experiment~3 using the \Galah\
			 data. \todo{The panels on the left show the absolute entries for each
			 factor load.} On the right we show the absolute fractional contributions
			 to each element, illustrating the loads that contribute most to each
			 element.}
    \label{fig:exp3-factor-loads}
\end{figure*}

@@

\begin{comment}
\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp3-abundance-counts.pdf}
	\caption{The number of abundance measurements in the second
		     \Galah\ data release \citep{Buder:2018} where \texttt{flag\_cannon = 0}
		     and \texttt{flag\_<element> = 0}. The black line indicates the number
		     of reported measurements for a single element, and the blue line
		     indicates the number of stars where all elements leftward (inclusive)
		     are reported. The dashed vertical line indicates the top 18 reported
		     elemental abundances, which we use for Experiment~4.}
    \label{fig:exp3-abundance-counts} 
\end{figure}
\end{comment}

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp3-\ExperimentHash-specific-scatter.pdf}
	\caption{Specific scatter $\sqrt{\specificvariance}$ remaining in the \Galah\ data \citep{Buder:2018}
			 in Experiment~3 after accounting for the contributions by all
			 latent factors.}
    \label{fig:exp3-specific-scatter}
\end{figure}

\begin{figure*}
	\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-latent-space.pdf}
	\caption{\todo{Caption}}
    \label{fig:exp3-latent-space}
\end{figure*}

\begin{figure*}
	\includegraphics[width=\textwidth]{experiments/exp3-\ExperimentHash-data-space.pdf}
	\caption{\todo{Caption}}
    \label{fig:exp3-data-space}
\end{figure*}
